---
title: "Explore Extracting PA Supplemental Homicde Data from Excel Files"
author: "Earl F Glynn<br>"
date: "<small>`r Sys.Date()`</small>"
output:
  html_document:
    code_download: true
    theme: cerulean
    toc: yes
    toc_depth:  3
    toc_float:
      collapsed:  yes
      smooth_scroll: yes
    number_sections: yes
    code_folding:  show
---

```{r setup, echo = FALSE}
# http://biostat.mc.vanderbilt.edu/wiki/Main/KnitrHtmlTemplate
require(Hmisc)    # provides knitrSet and other functions
knitrSet(lang = 'markdown',   # If using blogdown: knitrSet(lang='blogdown')
         fig.align = 'left',
         w = 6.5,
         h = 4.5,
         cache = FALSE)
```

`r hidingTOC(buttonLabel = "Outline")`

```{r startYourEngines, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA)

time.1 <- Sys.time()
```

# Setup {.tabset .tabset-fade .tabset-pills}

## {.active}

## Constants

```{r Constants}
FILE_PREFIX   <- "x"    # no spaces
```

```{r}
EXCEL_LIMIT <- 2^20
```

## Packages

```{r Packages}
library(tidyverse)
library(lubridate)   # mdy
```

Display formatting

```{r Display}
library(kableExtra)  # kable_styling
#library(IRdisplay)   # display_html (needed in Jupyter notebooks)
```

I/O

```{r IO}
library(readxl)      # read_xlsx
library(readr)       # write_csv
library(writexl)     # write_xlsx
```

 Graphics

```{r Graphics}
library(scales)      # comma, comma_format
```

## ggplot theme

[ggplot Themes Gallery](https://www.datanovia.com/en/blog/ggplot-themes-gallery/)

```{r ggplotTheme}
theme_set(theme_minimal() +

          theme(axis.text             = element_text(size = 10),
                axis.title            = element_text(size = 14),

                plot.caption          = element_text(hjust = c(0.0,1.0)),
                plot.caption.position = "plot",   # R 4.0.0

                plot.title.position   = "plot",
                title                 = element_text(size = 14))
           )

COLOR_BAR     <- "skyblue"
COLOR_OUTLINE <- "grey80"
```

## Helper functions

```{r Helpers}
Show <- function(data, caption="", bigMark="",
                 height = NULL, width = NULL, ...)
{
  data                                       %>%
  kable("html", caption=caption,
        format.args=list(big.mark=bigMark))  %>%
  kable_styling(bootstrap_options=c("striped", "bordered", "condensed"),
                position="left",
                full_width=FALSE, ...)       %>%
  scroll_box(height = height, width = width) #  %>%
# as.character()                             %>%      # Jupyter Only
# display_html()                                      # Jupyter Only
}
```

# Data Source 

`download` directory

1. Start on this page:  [Crime in Pennsylvania Online](https://www.ucr.pa.gov/PAUCRSPUBLIC/ReportsIndex/List)

2. Select [Annual SRS Summary Report](https://www.ucr.pa.gov/PAUCRSPUBLIC/SRSReport/AnnualSRSSummary)

3. Select Report Type:  Supplemental Homicide

4. Step single month by specifying `Start` and `End` the same month, e.g., Sep - 2022

5. Select:  Generate Report

6. Select floppy icon dropdown:  Excel

7. Save to file:  SRS_AnnualSummarySupplementalHomicide-YYYY-MM.xlsx

8. Repeat for 92 months:  2015-01 through 2022-09.


# Setup table column names

```{r}
race     <- c("RW", "RB", "RN", "RA", "RH", "RU")
sex      <- c("SM", "SF", "SU")
ethnic   <- c("EH", "EN", "EU")
```


```{r}
raceCodes <- t( outer(race, sex, FUN = paste0) ) %>% as.vector() 
raceCodes
```

```{r}
ethnicCodes <- t( outer(ethnic, sex, FUN = paste0) ) %>% as.vector() 
ethnicCodes
```

```{r}
columnCodes <- c("Age", raceCodes, ethnicCodes)
columnCodes
```

# Read File

```{r}
yyyymm <- "2022-08"
yyyymm
```

```{r}
filename <- paste0("download/SRS_AnnualSummarySupplementalHomicide-", yyyymm, ".xlsx")
d <- read_xlsx(filename)

dim(d)
```

```{r}
write_xlsx(d, paste0("rawdata/", yyyymm, "-Reformat.xlsx"))
```

# Extract data

```{r}
beginDate <- str_sub(d[6, 12], -10) %>% mdy()  # last 10 characters
beginDate
```

```{r}
endDate <- str_sub(d[7, 35], -10) %>% mdy()
endDate
```

```{r}
agency <- str_sub(d[13, 1], -3)
agency
```

# Victim

```{r}
section1 <- d[15, 1]
section1
```

### Properties

```{r}
columns <- which(!is.na(d[16, 1 : ncol(d)]))
columns
```

```{r}
t( d[16, columns] )
```

Verify property values as expected

```{r}
stopifnot( all(d[16, columns] == c("Race", "Ethnic Origin")) )
```

### Race / Ethnicity

```{r}
columns <- which(!is.na(d[17, 1 : ncol(d)]))
columns
```

```{r}
t(d[17, columns])
```

Verify race values as expected

```{r}
stopifnot( all(d[17, columns] == c("White", "Black /African American", "Native American / Alaskan Native",
                                   "Asian", "Native Hawaiian / Other Pacific Islander", "Unknown",
                                   "Hispanic", "Not Hispanic", "Unknown")) )
```

### Sex

```{r}
columns <- which(!is.na(d[18, 1 : ncol(d)]))
columns
```

```{r}
t( d[18, columns] )
```

Verify sex values as expected

```{r}
stopifnot( all( d[18, columns]  == rep(c("M", "F", "U"), length(columns) %/% 3)) )
```

### Victim table

```{r}
columns <- c(1, which(!is.na(d[18, 1 : ncol(d)])))
columns
```

```{r}
victim <- d[19:38, columns]
dim(victim)
```

```{r}
colnames(victim) <- columnCodes
```

```{r}
victim <- 
  victim  %>%
  mutate(across(starts_with(c("R", "E")), as.numeric) )
```


```{r}
victim <- 
  victim                                                 %>%
  mutate(across(starts_with(c("R", "E")), as.numeric) )  %>%
  mutate(Agency = agency,
         Table  = "Victim", 
         Begin  = beginDate, 
         End    = endDate)                               %>%
  relocate(Agency, Table, Begin, End)
```


```{r}
victim %>% Show()
```

# Offender

```{r}
section2 <- d[41, 1]
section2
```

### Properties

```{r}
columns <- which(!is.na(d[42, 1 : ncol(d)]))
columns
```

```{r}
d[42, columns]
```

verify property values as expected


```{r}
stopifnot( all(d[42, columns] == c("Race", "Ethnic Origin")) )
```

### Race / Ethnicity


```{r}
columns <- which(!is.na(d[43, 1 : ncol(d)]))
columns
```
```{r}
t(d[43, columns])
```

Verify race values as expected

```{r}
stopifnot( all(d[43, columns] == c("White", "Black /African American", "Native American / Alaskan Native",
                                   "Asian", "Native Hawaiian / Other Pacific Islander", "Unknown",
                                   "Hispanic", "Not Hispanic", "Unknown")) )
```

### Sex

```{r}
columns <- which(!is.na(d[44, 1 : ncol(d)]))
columns
```

```{r}
t( d[44, columns] )
```

Verify sex values as expected

```{r}
stopifnot( all( d[44, columns]  == rep(c("M", "F", "U"), length(columns) %/% 3)) )
```

### Offender table

```{r}
columns <- c(1, which(!is.na(d[44, 1 : ncol(d)])))
columns
```

```{r}
offender <- d[45:64, columns]
dim(offender)
```

```{r}
colnames(offender) <- columnCodes
```

```{r}
offender <- 
  offender                                               %>%
  mutate(across(starts_with(c("R", "E")), as.numeric) )  %>%
  mutate(Agency = agency,
         Table  = "Offender", 
         Begin  = beginDate, 
         End    = endDate)                               %>%
  relocate(Agency, Table, Begin, End)
```

```{r}
offender %>% Show()
```

# Final Extracted Table

```{r}
final <- bind_rows(victim, offender)
dim(final)
```

```{r}
write_xlsx(final, paste0("tables/", yyyymm, "-Victim-Offender-Data.xlsx"))
```

```{r}
final %>% Show()
```

# Epilog {.tabset .tabset-fade .tabset-pills}

## {.active}

## Session Info

```{r devtoolsSessionInfo}
devtools::session_info()
```

</div>

```{r epilogDeltaTime, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", sprintf("%.1f",
                        as.numeric(difftime(time.2,
                                            time.1, units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`

