---
title: "Extracting PA Supplemental Homicde Data from Excel Files"
author: "Earl F Glynn<br>"
date: "<small>`r Sys.Date()`</small>"
output:
  html_document:
    code_download: true
    theme: cerulean
    toc: yes
    toc_depth:  3
    toc_float:
      collapsed:  yes
      smooth_scroll: yes
    number_sections: yes
    code_folding:  show
---

```{r setup, echo = FALSE}
# http://biostat.mc.vanderbilt.edu/wiki/Main/KnitrHtmlTemplate
require(Hmisc)    # provides knitrSet and other functions
knitrSet(lang = 'markdown',   # If using blogdown: knitrSet(lang='blogdown')
         fig.align = 'left',
         w = 6.5,
         h = 4.5,
         cache = FALSE)
```

`r hidingTOC(buttonLabel = "Outline")`

```{r startYourEngines, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA)

time.1 <- Sys.time()
```

# Setup {.tabset .tabset-fade .tabset-pills}

## {.active}

## Constants

```{r Constants}
FILE_PREFIX   <- "x"    # no spaces
```

```{r}
EXCEL_LIMIT <- 2^20
```

## Packages

```{r Packages}
library(tidyverse)
library(lubridate)   # mdy
```

Display formatting

```{r Display}
library(kableExtra)  # kable_styling
#library(IRdisplay)   # display_html (needed in Jupyter notebooks)
```

I/O

```{r IO}
library(readxl)      # read_xlsx
library(readr)       # write_csv
library(writexl)     # write_xlsx
```

 Graphics

```{r Graphics}
library(scales)      # comma, comma_format
```

## ggplot theme

[ggplot Themes Gallery](https://www.datanovia.com/en/blog/ggplot-themes-gallery/)

```{r ggplotTheme}
theme_set(theme_minimal() +

          theme(axis.text             = element_text(size = 10),
                axis.title            = element_text(size = 14),

                plot.caption          = element_text(hjust = c(0.0,1.0)),
                plot.caption.position = "plot",   # R 4.0.0

                plot.title.position   = "plot",
                title                 = element_text(size = 14))
           )

COLOR_BAR     <- "skyblue"
COLOR_OUTLINE <- "grey80"
```

## Helper functions

```{r Helpers}
Show <- function(data, caption="", bigMark="",
                 height = NULL, width = NULL, ...)
{
  data                                       %>%
  kable("html", caption=caption,
        format.args=list(big.mark=bigMark))  %>%
  kable_styling(bootstrap_options=c("striped", "bordered", "condensed"),
                position="left",
                full_width=FALSE, ...)       %>%
  scroll_box(height = height, width = width) #  %>%
# as.character()                             %>%      # Jupyter Only
# display_html()                                      # Jupyter Only
}
```

# Data Source 

`download` directory

1. Start on this page:  [Crime in Pennsylvania Online](https://www.ucr.pa.gov/PAUCRSPUBLIC/ReportsIndex/List)

2. Select [Annual SRS Summary Report](https://www.ucr.pa.gov/PAUCRSPUBLIC/SRSReport/AnnualSRSSummary)

3. Select Report Type:  Supplemental Homicide

4. Step single month by specifying `Start` and `End` the same month, e.g., Sep - 2022

5. Select:  Generate Report

6. Select floppy icon dropdown:  Excel

7. Save to file:  SRS_AnnualSummarySupplementalHomicide-YYYY-MM.xlsx

8. Repeat for 92 months:  2015-01 through 2022-09.

See `00-Explore-Extraction.html` for detailed look at extracting the data.

A function will read and process each each Excel file.

# Processing function

Many redundant checks to verify the format is the same across all reports.

```{r}
processHomicideData <- function(yyyymm)
{

  # Setup table column names
  
  race     <- c("RW", "RB", "RN", "RA", "RH", "RU")
  sex      <- c("SM", "SF", "SU")
  ethnic   <- c("EH", "EN", "EU")
  
  raceCodes <- t( outer(race, sex, FUN = paste0) ) %>% as.vector() 
  
  ethnicCodes <- t( outer(ethnic, sex, FUN = paste0) ) %>% as.vector() 
  
  columnCodes <- c("Age", raceCodes, ethnicCodes)
  
  
  filename <- paste0("download/SRS_AnnualSummarySupplementalHomicide-", yyyymm, ".xlsx")
  d <- read_xlsx(filename)


  # save `rawdata` version
  write_xlsx(d, paste0("rawdata/", yyyymm, "-Reformat.xlsx"))


  # Extract data

  beginDate <- str_sub(d[6, 12], -10) %>% mdy()  # last 10 characters
  endDate   <- str_sub(d[7, 35], -10) %>% mdy()
  
  agency <- str_sub(d[13, 1], -3)

  # Victim data with verification checks
  vOK1 <- d[15, 1] == "Victim Age"
  
  columns <- which(!is.na(d[16, 1 : ncol(d)]))
  vOK2 <- all(d[16, columns] == c("Race", "Ethnic Origin")) 
  
  columns <- which(!is.na(d[17, 1 : ncol(d)]))
  vOK3 <- all(d[17, columns] == c("White", "Black /African American", "Native American / Alaskan Native",
                                     "Asian", "Native Hawaiian / Other Pacific Islander", "Unknown",
                                     "Hispanic", "Not Hispanic", "Unknown")) 
  
  columns <- which(!is.na(d[18, 1 : ncol(d)]))
  vOK4 <-  all( d[18, columns]  == rep(c("M", "F", "U"), length(columns) %/% 3)) 
  
  columns <- c(1, which(!is.na(d[18, 1 : ncol(d)])))
  victim <- d[19:38, columns]
  colnames(victim) <- columnCodes
  
  victim <- 
    victim                                                 %>%
    mutate(across(starts_with(c("R", "E")), as.numeric) )  %>%
    mutate(Agency = agency,
           Table  = "Victim", 
           Begin  = beginDate, 
           End    = endDate)                               %>%
    relocate(Agency, Table, Begin, End)

  # Offender data with verification checks
  oOK1 <- d[41, 1] == "Offender Age"
  
  columns <- which(!is.na(d[42, 1 : ncol(d)]))
  oOK2 <- all(d[42, columns] == c("Race", "Ethnic Origin")) 
  
  columns <- which(!is.na(d[43, 1 : ncol(d)]))
  oOK3 <- all(d[43, columns] == c("White", "Black /African American", "Native American / Alaskan Native",
                                     "Asian", "Native Hawaiian / Other Pacific Islander", "Unknown",
                                     "Hispanic", "Not Hispanic", "Unknown")) 
  
  columns <- which(!is.na(d[44, 1 : ncol(d)]))
  oOK4 <-  all( d[44, columns]  == rep(c("M", "F", "U"), length(columns) %/% 3)) 

  columns <- c(1, which(!is.na(d[44, 1 : ncol(d)])))
  offender <- d[45:64, columns]

  colnames(offender) <- columnCodes

  offender <- 
    offender                                               %>%
    mutate(across(starts_with(c("R", "E")), as.numeric) )  %>%
    mutate(Agency = agency,
           Table  = "Offender", 
           Begin  = beginDate, 
           End    = endDate)                               %>%
    relocate(Agency, Table, Begin, End)

  # Final Extracted Table
  final <- bind_rows(victim, offender)

  write_xlsx(final, paste0("tables/", yyyymm, "-Victim-Offender-Data.xlsx"))
  
  list(
        table      = final, 
        victimOK   = vOK1 && vOK2 && vOK3 && vOK4,
        offenderOK = oOK1 && oOK2 && oOK3 && oOK4
      )
}
```

# Test Function

```{r}
yyyymm <- "2022-08"
d <- processHomicideData(yyyymm)

str(d)
```

# Process all downloaded files

```{r}
yearMonthList <- 
  list.files(path = "download", glob2rx("*.xlsx")) %>% 
  str_sub(-12)  %>%
  str_sub(1, 7)

length(yearMonthList)
```

```{r}
yearMonthList
```

```{r}
finalTable <- NULL

for (i in 1:length(yearMonthList))
{
  cat(yearMonthList[i], "\n")
  result <- processHomicideData(yearMonthList[i])
  if (result$victimOK && result$offenderOK)
  {
    finalTable <- bind_rows(finalTable, result$table)
  } else
  {
    cat("   victim", result$victimOK, "offender", result$offenderOK, "\n")
  }
    
}

dim(finalTable)
```

## Explore and Verify

### `Offender` and `Victim table counts match

```{r}
finalTable     %>%
  count(Table) %>%
  Show()
```

## Same `Age` intervals alwaysw present

```{r}
finalTable    %>%
  count(Age)  %>%
  Show(height = "200px")
```

## Introduce `AgeIntervals`

Recode `Age` into fewer and more consistent intervals

```{r}
AgeIntervals <-
  tibble(
         Age = c( 
                 "UNDER 1", "01-04", "05-09",
                 "10-14", "15-16", "17-19",
                 "20-24", "25-29",
                 "30-34", "35-39",
                 "40-44", "45-49",
                 "50-54", "55-59",
                 "60-64", "65-69",
                 "70-74", "75 and UP",
                 "Unknown",
                 "Total"
                ),
         AgeInterval = c(
                          "00s",  "00s",  "00s",
                          "10s",  "10s",  "10s",
                          "20s",  "20s",
                          "30s",  "30s",
                          "40s",  "40s",
                          "50s",  "50s",
                          "60s",  "60s",
                          "70s+", "70s+",
                          "Unknown",
                          "Total"
                        )
        )

AgeIntervals %>% Show()
```

Add `AgeIntervals` to tibble

`group_by` by `AgeIntervals` for aggregationws instead of `Age` (but ignore `Total`)

```{r}
finalTable <- 
  finalTable %>%
  inner_join(AgeIntervals, by = "Age")  %>%
  relocate(AgeInterval, .after = Age)
```

# Aggregate all sexes by race and ethnic origin

```{r}
finalTable <-
  finalTable %>%
  mutate(
          RWhite    = RWSM +	RWSF + RWSU,
          RBlack    = RBSM	+ RBSF + RBSU,
          RNative   = RNSM	+ RNSF + RNSU,
          RAsian    = RASM	+ RASF + RASU,
          RHawaiian = RHSM	+ RHSF + RHSU,
          RUnknown  = RUSM	+ RUSF + RUSU,
          RTotal    = RWhite + RBlack + RNative + RAsian + RHawaiian + RUnknown,
        
          EHispanic    = EHSM	+ EHSF + EHSU,
          ENonHispanic = ENSM +	ENSF + ENSU,
          EUnknown     = EUSM	+ EUSF + EUSU,
          ETotal       = EHispanic + ENonHispanic + EUnknown
        )
```

```{r}
glimpse(finalTable)
```


```{r}
write_xlsx(finalTable, "Pennsylvania-Homicide-Data-Monthly-By-Age-Group.xlsx")
```

```{r}
mismatches <- (finalTable$RTotal == finalTable$ETotal) %>% as.integer()
table(mismatches, useNA = "ifany")
```

```{r}
write_xlsx(finalTable[mismatches == 0, ], "Pennsylvania-Homicide-Data-Monthly-By-Age-Group-Mismatches.xlsx")
```

```{r}
counts <- 
  finalTable            %>%
  group_by(Begin, End)  %>%
  count()               %>%
  ungroup()             %>%
  arrange(Begin)

counts %>%
  Show(height = "200px")
```

# Extract  Monthly `Total` lines

```{r}
monthlyTotals <-
  finalTable   %>%
  filter(Age == "Total")

nrow(monthlyTotals)
```

```{r}
write_xlsx(monthlyTotals, "Pennsylvania-Homicide-Data-Monthly-Totals.xlsx")
```

```{r}
mismatches <- (monthlyTotals$RTotal == monthlyTotals$ETotal) %>% as.integer()
table(mismatches, useNA = "ifany")
```

```{r}
write_xlsx(monthlyTotals[mismatches == 0, ], "Pennsylvania-Homicide-Data-Monthly-Totals-Mismatches.xlsx")
```

# Epilog {.tabset .tabset-fade .tabset-pills}

## {.active}

## Session Info

```{r devtoolsSessionInfo}
devtools::session_info()
```

</div>

```{r epilogDeltaTime, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", sprintf("%.1f",
                        as.numeric(difftime(time.2,
                                            time.1, units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`

